---
title: "Final Project RMD"
author: "Will Bannon and Kriss Vesu"
output: html_document
date: "2024-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```
Code should be collapsed/hidden for HTML, and R Notebooks or in a Code Appendix at the end for a PDF.<----
DO THIS



  We are trying to explore who is the best player in the history of the NBA according to statistical analysis of the two clear cut frontrunners: Lebron James and Michael Jordan. We used many data visualizations, summary tables, and a lot of data wrangling to analyze who might be the better player. 

  We retrieved our two datasets from Kaggle with the career statistics of Lebron James and Michael Jordan each in separate datasets. The data was originally compiled for the same purpose we intend to use them, however Kaggle only had a repository of raw data whereas we ran analysis upon said raw data.

ctrl-alt-i


```{r data wrangling}
lebron_data <- read.csv("C:/Users/willb/Desktop/desktopfolder/JordanLebronDataset/lebron_career.csv")
lebron_data$age <- substr(lebron_data$age, 1, 2)
write.csv(lebron_data, "modified_lebron_data.csv", row.names = FALSE)

jordan_data <- read.csv("C:/Users/willb/Desktop/desktopfolder/JordanLebronDataset/jordan_career.csv")
jordan_data$age <- substr(jordan_data$age, 1, 2)
write.csv(jordan_data, "modified_jordan_data.csv", row.names = FALSE)
#In this code we read in the data from our file directory.


lebron_data$ts = (lebron_data$fg + (0.5 * lebron_data$three)) / lebron_data$fga
jordan_data$ts = (jordan_data$fg + (0.5 * jordan_data$three)) / jordan_data$fga
#In this code we created a new column to our dataset called True Shooting Percentage which measures a players scoring efficiency more aptly than the other columns provided to us in the dataset. The formula for True Shooting Percentage is calculated using field goals made, three point field goals made, and field goals attempted.

lebron_data$name = "LBJ"
jordan_data$name = "MJ"

new_table <- rbind(lebron_data, jordan_data)
View(new_table)
#In this code we combined the two data sets into one singular table so that we might begin data manipulation with more ease. The new table can be viewed here.
```


DISCUSS THE GRAPH
```{r}
ggplot() +
  geom_boxplot(data = new_table, aes(x = age, y = pts, color = name)) +
  scale_color_manual(values = c("red", "blue")) +
  labs(title = "Points by each GOAT",
       x = "Age",
       y = "Points per Game")
```

